################################################################################################
 R Project on Linear Regression and Residual Analysis                                 

 Last Revision: 04/18/2019

 Author(s):
 Lee Ping Tay - joylp.tay@gmail.com

 
 Description:

 Source of Data: 
 The Integrated Postsecondary Education Data System (IPEDS) by the National Center for 
 Education Statistics (NCES) in United States.
 (https://nces.ed.gov/ipeds/datacenter/DataFiles.aspx)

 Goal:
 This R script is used to perform exploratory data analysis and data visualization on
 higher eduacation data. Bar graphs on most expensive and least expensive colleges 
 were created. Linear Regression and Residuals Analysis were conducted to identify
 associations between Tuition Fees and Admission Rate.

 Contents: 
 Libraries and Environment
 Data Import and Preprocessing
 Data Manipulation (Merge dataframes)
 Exploratory Data Analysis / Data Visualization
 Linear Regression
 Residuals analysis 
 
###############################################################################################
###############################################################################################
 Libraries and Environment
###############################################################################################

```{r }
library(tidyverse)
library(RColorBrewer)
```

###############################################################################################
 Data Import and Data Manipulation
###############################################################################################
 Import cvs file

```{r }
df_ipeds_data <- as.data.frame(read.csv("hd2017.csv"))
```

 Filter data on ipeds information

```{r }
df_ipeds <- df_ipeds_data %>%
               select(UNITID, INSTNM, ADDR, CITY, STABBR, ZIP, CSA, COUNTYCD, 
                      COUNTYNM, LONGITUD, LATITUDE, CONTROL)
```

dim(df_ipeds)   7153   12
 Import Student charges for academic year programs

```{r }
tuitionfee <- as.data.frame(read_csv("ic2017_ay.csv"))
```

 Filter data on in-state tuition and fees 2017-18 (CHG2AY3) and
 out-of-state tuition and fees 2017-18 (CHG3AY3)

```{r }
df_tuitionfee  <-  tuitionfee %>%
                   select(UNITID, CHG2AY3, CHG3AY3) 
```

dim(df_tuitionfee)    4281    3
 Import Admissions and Test Scores

```{r }
adm <- as.data.frame(read_csv("adm2017.csv"))
```

 Filter data on ADMSSN (total admissions), APPLCN (total applicants), ENRLT (total enrolled)

```{r }
df_admission  <- adm %>%
                  select(UNITID, ADMSSN, APPLCN , ENRLT) 
```

dim(df_admission)  2075    4
 import data on financial information

```{r }
cscore_all <- read.csv('Most-Recent-Cohorts-All-Data-Elements.csv')
```

dim(cscore_all)   7175 1899
 filter data on debts, earnings, and tuition fee (in-state and out-of-state)

```{r }
fin_data <- cscore_all %>%
            select(UNITID, GRAD_DEBT_MDN_SUPP, MD_EARN_WNE_P10, TUITIONFEE_IN, TUITIONFEE_OUT) %>%
            mutate(MD_EARN_WNE_P10 = round(as.numeric(as.character(MD_EARN_WNE_P10))),
                   GRAD_DEBT_MDN_SUPP = round(as.numeric(as.character(GRAD_DEBT_MDN_SUPP))),
                   TUITIONFEE_IN = as.numeric(as.character(TUITIONFEE_IN)),
                   TUITIONFEE_OUT = as.numeric(as.character(TUITIONFEE_OUT)))
```

 Import data on regions from github 

```{r }
regions <- read.csv("https://raw.githubusercontent.com/cphalpert/census-regions/master/us%20census%20bureau%20regions%20and%20divisions.csv")
```

###############################################################################################
 Data Manipulation (Merge Dataframes)
###############################################################################################
1. Merge the data frame df_ipeds_main with the df_tuitiondfee data frame on UNITID

```{r }
df_ipeds_tuition <- merge(df_ipeds, df_tuitionfee, by = "UNITID", all.x = TRUE)
```

2. Merge the data frame IPEDS_merge_price_2017 with the adm_enroll2017 data frame on UNITID

```{r }
df_ipeds_tuiadm <- merge(df_ipeds_tuition, df_admission, by = "UNITID", all.x = TRUE)
```

2. Merge the data frame IPEDS_merge_price_2017 with the adm_enroll2017 data frame on UNITID

```{r }
df_ipeds_fin <- merge(df_ipeds_tuiadm, fin_data, by = "UNITID", all.x = TRUE)
```

4. Merge the data frame IPEDS_merge_adm_2017 with US Region(4) CW on STABBR

```{r }
df_ipeds_main <- merge(df_ipeds_fin, regions, by.x = "STABBR", by.y= "State.Code", all.x = TRUE)
```

dim(df_ipeds_main)   7153   24
###############################################################################################
 Exploratory Data Analysis / Data Visualization
###############################################################################################
class(df_ipeds_main$CHG2AY3)    "character"
 Change the class of tuition fee data and filter data on public and private not-for-profit 
 colleges

```{r }
df_ipeds_main1 <- df_ipeds_main %>%
                  mutate(CHG2AY3 = as.numeric(CHG2AY3),
                         CHG3AY3 = as.numeric(CHG3AY3)) %>%
                  filter(!is.na(Region) & !is.na(CHG2AY3) & CONTROL !=3)
```

dim(df_ipeds_main1)   3011   24
 Filter NA and calculate admission rate

```{r }
df_ipeds_adm <- df_ipeds_main1 %>%
                filter(!is.na(ADMSSN) & !is.na(ENRLT)) %>%
                mutate(adm_rate = round((ADMSSN/APPLCN)*100))
```

dim(df_ipeds_adm)   1705   25
 Average admission rate
mean(df_ipeds_adm$adm_rate)   67.25
 Top 10 Most Expensive Colleges in the U.S. in 2018

```{r }
top10_high_tuitionfee <- df_ipeds_main1 %>%
                         mutate(CHG2AY3 = as.numeric(CHG2AY3) ) %>%
                         filter(!is.na(CHG2AY3)) %>%
                         select(INSTNM, Region, CHG2AY3) %>%
                         top_n(n=10, wt =CHG2AY3) %>% 
                         arrange(-CHG2AY3) %>%
                         mutate(INSTNM=as.factor(INSTNM)) %>%
                         rename('School' = INSTNM,
                                'In_State_TuitionFees' = CHG2AY3)
```

 Top 10 Least Expensive Colleges in the U.S. in 2018

```{r }
top10_low_tuitionfee <- df_ipeds_main1 %>%
                        mutate(CHG2AY3 = as.numeric(CHG2AY3) ) %>%
                        filter(!is.na(CHG2AY3)) %>%
                        select(INSTNM, Region, CHG2AY3) %>%
                        top_n(n=10, wt = -CHG2AY3) %>% 
                        arrange(CHG2AY3) %>%
                        mutate(INSTNM=as.factor(INSTNM)) %>%
                        rename('School' = INSTNM,
                               'In_State_TuitionFees' = CHG2AY3)
```

 Set common colors for regions

```{r }
dd <- union(top10_high_tuitionfee$Region,top10_low_tuitionfee$Region)
dd.col <- brewer.pal(n=length(dd), name="Set3")
names(dd.col)  <- dd
```

 Bar Graph for Top 10 Most Expensive Colleges in the U.S. in 2018

```{r }
ggplot(top10_high_tuitionfee, aes(x=reorder(School, In_State_TuitionFees), 
                                  y=In_State_TuitionFees, fill=Region)) +
  theme(panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  geom_col() + coord_flip() +
  scale_fill_manual("Legend", values = dd.col) +
  labs(title = "Top 10 Most Expensive Colleges in the U.S. in 2018") +
  labs(x = "Name of College", y="In-State Tuition and Fees") +
  labs(caption="Source: IPEDS")
```

 Bar Graph for Top 10 Least Expensive Colleges in the U.S. in 2018

```{r }
ggplot(top10_low_tuitionfee, aes(x=reorder(School, In_State_TuitionFees), 
                                y=In_State_TuitionFees, fill =Region)) +
  theme(panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  geom_col() + coord_flip() +
  scale_fill_manual("Legend", values = dd.col) +
  labs(title = "Top 10 Least Expensive Colleges in the U.S. in 2018") +
  labs(x = "Name of College", y="In-State Tuition and Fees") +
  labs(caption="Source: IPEDS")
```

###############################################################################################
 Linear Regression
###############################################################################################
 fit the model

```{r }
plot(df_ipeds_adm$CHG2AY3 ~ df_ipeds_adm$adm_rate , xlab="Admission Rate", 
     ylab="In-State Tuition and Fees", main="Tuition and Fees vs Admission Rate", 
     col="blue")

tuitionfit_adm <- lm(formula = df_ipeds_adm$CHG2AY3 ~ df_ipeds_adm$adm_rate)
abline(tuitionfit_adm, col="red")
```

 report the result

```{r }
summary(tuitionfit_adm)
```

 r squared
summary(tuitionfit_adm)$r.squared   0.09
 coefficient of determination R^2 = 0.09
 9% of SST, squared variation in dependent variable, admission rate, 
 is explained by the linear regression equation to predict the in-state tuition and fees,
 the rest is error.
 print p-value

```{r }
summary(tuitionfit_adm)$coef[,"Pr(>|t|)",drop=F]
```

                            Pr(>|t|)
 (Intercept)             2.467593e-185
 df_ipeds_adm$adm_rate   1.045081e-36
 print coefficients for the linear model

```{r }
coefficients(tuitionfit_adm)
```

  (Intercept)      df_ipeds_adm$adm_rate 
   37151.6746             -206.9259 
 print the correlation coefficient

```{r }
cor(df_ipeds_adm$adm_rate, df_ipeds_adm$CHG2AY3)
```

 correlation coefficient r = -0.2996
 the variables have negative linear relationship
 95% confident that the coefficients are in these ranges

```{r }
confint(tuitionfit_adm, level=0.95)
```

                            2.5 %       97.5 %
  (Intercept)           34945.7030   39357.6461
  df_ipeds_adm$adm_rate  -238.2401    -175.6118
######################################################
The final linear model is of the form:
  y = 37151.6746  - 206.9259  * x
#######################################################
 ggplot (Earnings vs GPA Award) with colors and abline

```{r }
ggplot(df_ipeds_adm, aes(x = df_ipeds_adm$adm_rate, y = df_ipeds_adm$CHG2AY3,
                               color = Region)) + geom_point() + 
  theme(panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  ggtitle("College Tuition and Fees versus Admission Rate") + 
  xlab("Admission Rate") + ylab("In-State Tuition and Fees") +
 geom_abline(aes(slope= -238.2401, intercept = 34945.7030), size=1, color="orange") 
```

###############################################################################################
 Residuals analysis 
###############################################################################################
  diagnostic plot
  Plot the model information
  par(mfrow = c(2, 2))

```{r }
plot(tuitionfit_adm, cex.lab = 1.5, cex.axis = 1.5, col="blue")
```

We must test these three assumptions after calculating the coefficients in the linear 
regression equation using the residuals:
  
Residuals or errors are normally distributed
Residuals or errors have a constant variance
Residuals or errors are independent of one another (not relevant to our scenario)
We use residual analysis to determine whether or not these conditions have been met.
   Interpretation of the diagnostic plot
1. Residuals vs Fitted - shows if residuals have non-linear patterns.
   We examine if the residuals exhibit homoscedasticity or approximately constant 
   variance across the predicted values of the dependent variable. In this plot,
   Residuals are almost symmetrically distributed before observations pass 30000. 
   But there is no distinctive pattern.
2. Normal Q-Q - shows if residuals are normally distributed.
   Residuals lined well on the dotted line except a few observations at the two ends.
3. Scale-Location - shows if residuals are spread equally along the ranges of predictors.
   The line is almost horizontal and residuals appear randomly spread.
4. Residuals vs Leverage - find influential cases if any.
   Cook's distance" measure the influence of each observation on the regression coefficients.
   There is no influential case as we can barely see Cook’s distance lines (a red dashed line)
   because almost all cases are well inside of the Cook’s distance lines.
 obtain predicted and residual values (differences between each data point and linear model)

```{r }
df_ipeds_adm  <- df_ipeds_adm %>% 
                  mutate(predicted = predict(tuitionfit_adm),
                         residuals = residuals(tuitionfit_adm))
```

 df_ipeds_adm$predicted <-  predict(tuitionfit_adm)
 df_ipeds_adm$residuals <-  residuals(tuitionfit_adm)

```{r }
df_ipeds_adm %>% select(CHG2AY3, predicted, residuals) %>% head()
```

 Normality
 visually test the normality of the residuals

```{r }
hist(df_ipeds_adm$residuals, col="skyblue", xlab="Residuals", main="Distribution of Residuals")
```

 distribution is slightly skewed to the right
 Scatterplot of Residuals versus Admission Rate

```{r }
plot(df_ipeds_adm$residuals, df_ipeds_adm$adm_rate, col="skyblue",
     xlab="residuals", ylab="Admission Rate", main="Residuals versus Admission Rate")
```

 We can also use errors to look at Normality

```{r }
df_ipeds_adm <- df_ipeds_adm %>% 
                 mutate(err=(CHG2AY3 - predicted) / CHG2AY3)

plot(df_ipeds_adm$err, xlab = "Difference between actual and predicted tuition and fees", 
     ylab="percentage change", main="Errors between Actual and Predicted Tuition and Fees", 
     col="blue")
```

 Homoscedasticity or heteroscedastic 
 visually test the constant variance assumption

```{r }
tuitionfitted_adm <- fitted(tuitionfit_adm)

plot(tuitionfitted_adm, df_ipeds_adm$residuals, xlab="residuals", 
     ylab="Predicted In-state Tuition and Fees", col="blue")
```

 The graph shows a slightly different vertical height across the point cloud.
 Overall, residuals have constant variance across the predicted dependent variable,
 Note: From the plots, we noticed a few admission rates with extreme value if compare to average
       admission rate of 68% in the dataset
 136  - 2%  (Minerva Schools, CA)
 1078 - 22% (United States Merchant Marine Academy, NY)
 1290 - 3% (Curtis Institute of Music, PA)
ggplot on predicted tuition and fees vs admission rate

```{r }
p1 <- ggplot(df_ipeds_adm, aes(x=adm_rate, y=CHG2AY3)) +
```

      geom_smooth(method = "lm", se = FALSE, color = "lightgrey") +  add the regression slope. adjust alpha to fade out connection lines
      geom_segment(aes(xend=adm_rate, yend=predicted), alpha=.2) +  connect actual data point with predicted value
      geom_point(aes(color=abs(residuals))) +  color mapped to abs(residuals)
      scale_color_gradient2(low="blue", mid="white", high="red") +  color to use. highlight residuals(abs) made more red on actual values
      guides(color=FALSE) +   color legend removed
      geom_point(aes(y=predicted), shape=1) +  add the predicted value
      theme_bw() +   add theme

```{r }
      ggtitle("Predicted College Tuition and Fees versus Admission Rate") + 
      xlab("Admission Rate") + ylab("In-State Tuition and Fees")
p1

ggsave(filename = "predict_tuition_adm.png", p1, width = 7, height = 4, dpi = 300, 
       units = "in", device='png')
```

 There is more red for extreme values of gpa award where the actual values are
 greater than what is being predicted.
